Cascaded Approximation of Gaussian Blur

### 2015-05-15 article.tex
Add third section

\title{Cascaded Approximation of Gaussian Blur}
\author{Vabishchevich Nikolay}
\date{May 14, 2015}
\maketitle

@@

In this article I'm using two different sets of grid point positions. The first one is grid point at
every integer, $F^+_k$ (I call it even and mark with ``$+$''). The second one is grid points positioned
halfway between adjacent integers, $F^-_{k+1/2}$ (I call it odd and mark with ``$-$'').
Correspondingly there are two types of convolution filters: even
\begin{equation}\label{conv+}
    R^\pm_p = \sum_{k\in Z} F^\pm_{p+k}K^+_k,
\end{equation}
and odd
\begin{equation}\label{conv-}
    R^\mp_p = \sum_{k\in Z} F^\pm_{p+k+\frac12}K^-_{k+\frac12},
\end{equation}
where $K^\pm_p$ is filter kernel. Even filter doesn't change source oddness but odd filter do (hence
the name). Therefore our target problem is to effectively calculate approximation of
\begin{equation}
    R^\pm_p = \sum_{k\in Z} F^\pm_{p+k}G_k,
\end{equation}
where $F^\pm_p$ is the source function and $G_k$ is even gaussian kernel,
\begin{equation}\label{gauss_n}
    G_k = \frac1{\sqrt{2\pi\sigma^2}}\exp\left\{-\frac{k^2}{2\sigma^2}\right\},
\end{equation}
especially for large standard deviation $\sigma$.

@@

Going back to our gaussian filter (\ref{gauss_n}), its Fourier transform is another gaussian function
\begin{equation}\label{gauss}
    G^+(\varphi) \approx \exp\left\{-\frac{\sigma^2\varphi^2}2\right\}.
\end{equation}
That function has two different regions. The first is almost zero at high frequency and the second
is bell-like shape at low frequency. Correspondingly the task of approximation can be divided into
two subtasks described in the following sections.

@@

>>>>

\title{Cascaded Approximation of Gaussian Blur}
\author{Vabishchevich Nikolay}
\date{May 15, 2015}
\maketitle

@@

In this article I'm using two different sets of grid point positions. The first one is grid point at
every integer, $F^+_k$ (I call it even and mark with ``$+$''). The second one is grid points positioned
halfway between adjacent integers, $F^-_{k+1/2}$ (I call it odd and mark with ``$-$'').
Correspondingly there are two types of convolution filters: even
\begin{equation}\label{conv+}
    R^\pm_p = \sum_{k\in Z} F^\pm_{p+k}K^+_k,
\end{equation}
and odd
\begin{equation}\label{conv-}
    R^\mp_p = \sum_{k\in Z} F^\pm_{p+k+\frac12}K^-_{k+\frac12},
\end{equation}
where $K^\pm_p$ is filter kernel. Even filter doesn't change source oddness but odd filter do (hence
the name). Therefore our target problem is to efficiently calculate approximation of
\begin{equation}
    R^\pm_p = \sum_{k\in Z} F^\pm_{p+k}G_k,
\end{equation}
where $F^\pm_p$ is the source function and $G_k$ is even gaussian kernel,
\begin{equation}\label{gamma}
    G_k = \Gamma(\sigma^2;\, k) \equiv
        \frac1{\sqrt{2\pi\sigma^2}}\exp\left\{-\frac{k^2}{2\sigma^2}\right\},
\end{equation}
especially for large standard deviation $\sigma$.

@@

Going back to our gaussian filter (\ref{gamma}), its Fourier transform is another gaussian function
\begin{equation}\label{gauss}
    G(\varphi) \approx \exp\left\{-\frac{\sigma^2\varphi^2}2\right\}.
\end{equation}
That function has two different regions. The first is almost zero at high frequency and the second
is bell-like shape at low frequency. Correspondingly the task of approximation can be divided into
two subtasks described in the following sections.

@@

\section{Main low-frequency filter}

For last task of approximating central gaussian shape I've decided to use generic 9-tap filter as
the main workhorse. Under that name I imply even symmetric conservative convolutional filter of the
following form
\begin{equation}
    K_k = \left\{
        \begin{array}{ll}
            1 - \sum_{i=1}^4 C_i,& k = 0;\\
            \frac12 C_i,& k = m_i;\\
            0,& \text{otherwise};
        \end{array}
    \right.
\end{equation}
where $m_i$, $i=1, 2, 3, 4$, are sampling positions ($0 < m_1 < m_2 < m_3 < m_4$) and $C_i$ are
coefficients. Fourier transform of this filter kernel is
\begin{equation}
    K(\varphi) = 1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)].
\end{equation}
Further research shows that this filter isn't enough as central filter so I combine it with
additional stage of even binomial filter $\hat B^{2l}$ ($B^0$ mean ``do nothing''). Therefore
complete approximation of gaussian blur take the form of
\begin{equation}
    \hat G_{\text{apx}} = \hat Q^n\hat K\hat B^{2l}\hat P^n.
\end{equation}
Disregarding frequency crosstalk, diagonal part of that operator in Fourier space is
\begin{equation}
    G_{\text{apx}}(\varphi) = \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(2^nm_i\varphi)]\right\}
        \cos^{2l}(2^{n-1}\varphi) \prod_{k=0}^{n-1} \cos^{10}(2^{k-1}\varphi).
\end{equation}
Outside of the region $[-\pi/2^n, \pi/2^n]$ that should be already near zero, so we focus only on
center
\begin{equation}
    G_{\text{apx}}\left(\frac\varphi{2^n}\right) =
        \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right\}
        \cos^{2l}\frac\varphi2 \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}},
\end{equation}
where that expression should be near
\begin{equation}
    G\left(\frac\varphi{2^n}\right) = \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\}.
\end{equation}

One of the possible ways to find good kernel coefficients $C_i$ is to solve least squares problem.
Denote
\begin{equation}
    p(\varphi) \equiv \cos^{2l}\frac\varphi2 \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}}
\end{equation}
and
\begin{equation}
    g(\varphi) \equiv \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\},
\end{equation}
then the problem is to minimize
\begin{equation*}
    \lambda = \int_{-\pi}^\pi \left[G\left(\frac\varphi{2^n}\right) -
        G_{\text{apx}}\left(\frac\varphi{2^n}\right)\right]^2 w(\varphi) d\varphi =
\end{equation*}
\begin{equation*}
    \int_{-\pi}^\pi \left[g(\varphi) -
        \left\{1 - \sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right\}p(\varphi)\right]^2
        w(\varphi) d\varphi =
\end{equation*}
\begin{equation}
    \int_{-\pi}^\pi \left[g(\varphi) - p(\varphi) +
        p(\varphi)\sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right]^2 w(\varphi) d\varphi,
\end{equation}
where $w(\varphi) \approx 1$ is additional weight. As usual, calculating partial derivatives
\begin{equation*}
    0 = \frac12\frac{\partial\lambda}{\partial C_k} =
    \int_{-\pi}^\pi \left[g(\varphi) - p(\varphi) +
        p(\varphi)\sum_{i=1}^4 C_i[1 - \cos(m_i\varphi)]\right]
        p(\varphi)w(\varphi) [1 - \cos(m_k\varphi)] d\varphi =
\end{equation*}
\begin{equation}
    \int_{-\pi}^\pi [g(\varphi) - p(\varphi)]p(\varphi)w(\varphi) [1 - \cos(m_k\varphi)] d\varphi +
    \sum_{i=1}^4 C_i\int_{-\pi}^\pi p^2(\varphi)w(\varphi)
        [1 - \cos(m_i\varphi)][1 - \cos(m_k\varphi)] d\varphi.
\end{equation}
After defining
\begin{equation}
    U_k \equiv \frac1{2\pi}\int_{-\pi}^\pi p^2(\varphi)w(\varphi)e^{-ip\varphi}d\varphi
\end{equation}
and
\begin{equation}
    V_k \equiv \frac1{2\pi}\int_{-\pi}^\pi g(\varphi)p(\varphi)w(\varphi)e^{-ip\varphi}d\varphi
\end{equation}
we can get
\begin{equation}
    0 = \frac1{2\pi}\frac{\partial\lambda}{\partial C_k} = 2[V_0 - V_{m_k} - U_0 + U_{m_k}] +
        \sum_{i=1}^4 C_i[2U_0 - 2U_{m_i} - 2U_{m_k} + U_{m_i+m_k} + U_{|m_i-m_k|}].
\end{equation}
Resulting symmetric system of four linear equations
\begin{equation}
    \sum_{i=1}^4 C_i[2U_0 - 2U_{m_i} - 2U_{m_k} + U_{m_i+m_k} + U_{|m_i-m_k|}] =
        2[U_0 - U_{m_k} - V_0 + V_{m_k}],
\end{equation}
can be solved through standard methods of linear algebra.

The real problem here is to efficiently calculate $U_k$ and $V_k$. One method is to use crude
approximation
\begin{equation}
    p(\varphi) \approx \exp\left\{-\frac{\alpha\varphi^2}2\right\},
\end{equation}
where
\begin{equation}\label{alpha}
    \alpha = \frac l2 + \frac56(1 - 4^{-n}).
\end{equation}
In that case
\begin{equation}
    U_k \approx \frac1{2\pi}\int_{-\pi}^\pi
        \exp\left\{-\alpha\varphi^2\right\}e^{-ip\varphi}d\varphi \approx \Gamma(2\alpha;\, k),
\end{equation}
\begin{equation}\label{V-gauss}
    V_k \approx \frac1{2\pi}\int_{-\pi}^\pi
        \exp\left\{-\left[\frac{\sigma^2}{4^n} + \alpha\right]\frac{\varphi^2}2\right\}
        e^{-ip\varphi}d\varphi \approx \Gamma(\sigma^24^{-n} + \alpha;\, k),
\end{equation}
where $\Gamma$ is normalized gaussian function defined in (\ref{gamma}).

Much better method is to choose $w(\varphi)$ such that
\begin{equation}
    p(\varphi)w(\varphi) = \exp\left\{-\frac{\alpha\varphi^2}2\right\}
\end{equation}
holds exactly. In that case (\ref{V-gauss}) holds exactly too and for $U_k$ we
have
\begin{equation}
    U_k = \frac1{2\pi}\int_{-\pi}^\pi \exp\left\{-\frac{\alpha\varphi^2}2\right\}
        p(\varphi)e^{-ip\varphi}d\varphi =
    \frac1{2\pi}\int_{-\pi}^\pi \exp\left\{-\frac{\alpha\varphi^2}2\right\}
        B^{2l}(\varphi)D(\varphi)e^{-ip\varphi}d\varphi,
\end{equation}
where $D$ is effective rescale filter kernel with
\begin{equation}
    D(\varphi) \equiv \prod_{k=1}^n \cos^{10}\frac\varphi{2^{k+1}}.
\end{equation}
Therefore $U_k$ can be calculated in the following way: construct gaussian function
$\Gamma(\alpha;\, k)$, then apply filter $\hat D$, than apply binomial filter of order $2l$.
Alternatively that can be written in operator notation as
\begin{equation}
    \hat U = \hat B^{2l}\hat D\,\hat\Gamma(\alpha;\, \cdot).
\end{equation}

And the third method is to use $w(\varphi) = 1$ and calculate all precisely. In that case
\begin{equation}
    U_k = \frac1{2\pi}\int_{-\pi}^\pi B^{4l}(\varphi)D^2(\varphi)e^{-ip\varphi}d\varphi,
\end{equation}
\begin{equation}
    V_k = \frac1{2\pi}\int_{-\pi}^\pi
        \exp\left\{-\frac{\sigma^2\varphi^2}{2\cdot4^n}\right\}
        B^{2l}(\varphi)D(\varphi)e^{-ip\varphi}d\varphi,
\end{equation}
or alternatively
\begin{align}
    \hat U &= \hat B^{4l}\hat D^2,\\
    \hat V &= \hat B^{2l}\hat D\,\hat\Gamma(\sigma^24^{-n};\, \cdot).
\end{align}

\begin{figure}[t]
    \centering
    \begin{tikzpicture}
        \begin{axis} [ spectrum, width = 8cm, height = 8cm,
                name = plot1, title = { $D(\varphi)$, $n = 1$ } ]
            \addplot [ sharp plot, line width = 1pt, blue ] gnuplot {
                cos(x*pi/4)^10
            };
            \addplot [ sharp plot, line width = 1pt, red ] gnuplot {
                (381024 + 181440 * 2*cos(x*pi) + 15120 * 2*cos(2*x*pi)) / 774144
            };
        \end{axis}
        \begin{axis} [ spectrum, width = 8cm, height = 8cm,
                name = plot2, title = { $D(\varphi)$, $n = 2$ },
                at = { (plot1.right of east) }, anchor = left of west, xshift = 0.5cm ]
            \addplot [ sharp plot, line width = 1pt, blue ] gnuplot {
                cos(x*pi/4)^10 * cos(x*pi/8)^10
            };
            \addplot [ sharp plot, line width = 1pt, red ] gnuplot {
                (21981456 + 11993940 * 2*cos(x*pi) +
                1746360 * 2*cos(2*x*pi) + 41580 * 2*cos(3*x*pi)) / 49545216
            };
        \end{axis}
    \end{tikzpicture}
    \caption{Approximation of rescale kernel $D$ (exact curve showed as blue).}
    \label{D}
\end{figure}

Third method is usually slightly better than second and much better then first, but precise methods
require good approximation of effective rescale filter $D$. Experiments shows that tight 7-tap
approximation
\begin{equation}
    D(\varphi) \approx D_0 + 2\sum_{k=1}^3 D_k\cos(k\varphi)
\end{equation}
is quite good. Coefficients here should be chosen in such a way that equation is exact up to 6
term of Taylor series around $\varphi = 0$. Corresponding problem can be solved analytically
and give the following results
\newcommand\0{\phantom{0}}
\begin{align}
    D_0 &= (   5204 +   2520\cdot4^{-n} +   1092\cdot16^{-n} +   3280\cdot64^{-n}) / 12096,\\
    D_1 &= (   2943 - \0 210\cdot4^{-n} - \0 273\cdot16^{-n} -   2460\cdot64^{-n}) / 12096,\\
    D_2 &= (\0  486 - \0 924\cdot4^{-n} - \0 546\cdot16^{-n} + \0 984\cdot64^{-n}) / 12096,\\
    D_3 &= (\0\0 17 - \0 126\cdot4^{-n} + \0 273\cdot16^{-n} - \0 164\cdot64^{-n}) / 12096.
\end{align}
Fig.~\ref{D} shows that such approximation noticeably differ from exact curve only at region near
$\varphi = \pm\pi$ where the target gaussian is near zero anyway. Accurate approximation of
effective rescale filter kernel concludes calculation of the main filter coefficients, and in the
following section I study the task of choosing remaining parametrs $n$, $l$ and $m_i$.
